% FORMALISM OVERVIEW
%
% PURPOSE: Define a logic-agnostic formal language for expressing embedded/avionics
% software requirements and test cases, so that coverage gaps can be detected
% mechanically. The ten Embraer requirements in req/ are the concrete test bed.
%
% AUTHORITATIVE REFERENCE: ../abstract-final.md (8-section spec).
% This file is the LaTeX rendering of that spec.
%
% KEY DESIGN PRINCIPLES:
%
% 1. LAYERS
%    Primitives (T, V) -> Entities & Dictionary -> Valuations & Traces ->
%    Expressions & Predicates -> Operations -> Temporal Constructs -> Requirements
%    -> Test Cases & Coverage
%
% 2. EXPRESSIONS ARE EVALUATED AT AMBIENT (rho, t)
%    Every expression is implicitly a function of trace rho and current time t.
%    e.g. prev(e) means Val_rho(e, t-1) where t is supplied by the surrounding
%    temporal operator, NOT by the operation call. This means operations do NOT
%    take t as an explicit argument -- t is implicit.
%
% 3. OPERATIONS DEFINE POINT-IN-TIME EFFECT PREDICATES
%    An operation such as calculate(e, expr) produces a predicate that holds at
%    the *current* t. The temporal extent (always, eventually, from t onwards)
%    is specified at the requirement level via temporal operators (always, triggered,
%    etc.), not inside the operation definition.
%    Example: always(calculate(e, 1.2 * prev(e))) expresses the recurrence
%    e(k) = 1.2 * e(k-1) for all k -- the always handles "for all time".
%
% 4. THE DICTIONARY IS OPEN
%    Entity types (Signal, Storage, Event, Channel, State, Value, Mapping, Abstract)
%    each declare: semantics, applicable operations, and admitted modifiers.
%    New types are added by extending the dictionary, not by changing the core.
%
% 5. COVERAGE IS DERIVED FROM REQUIREMENT STRUCTURE
%    MC/DC for ALL_OF conditions, per-element for ANY_OF, boundary tests for time
%    bounds, negative cases for guards, ordering violation tests for strict sequences.

% Requirements for this docuement:
% - it needs to be written in simple wording, dropping any fluff

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{stmaryrd}
\usepackage[margin=2.5cm]{geometry}
\usepackage[normalem]{ulem}
\usepackage{hyperref}
\usepackage{nameref}
\usepackage{etoolbox}
\usepackage{xcolor}
\newcommand{\todo}[1]{\textcolor{red}{TODO: [#1]}}

\newtheorem{definition}{Definition}
\AtEndEnvironment{definition}{\vspace{0.5em}}

\title{PP1 - Evaluating Test Coverage Against Software Requirements -- Formalism}
\author{  }
\date{April 2026}

\begin{document}

\maketitle

\tableofcontents
\newpage

\include{formalism}

\newpage

\section{Design Decisions}

\begin{itemize}
    \item Retaining the entity type as an explicit field is justified because it allows us to map each type to its own set of applicable modifiers and to its own set of applicable operations. \todo{This might be outdated now.}
\end{itemize}

\section{Current Limitations}

\begin{itemize}
    \item Modifiers need to be parameterized. \todo{This is maybe done now? Need to verify.}
    \item \textbf{Partiality of time functions}: several functions are explicitly partial -- $\mathit{Prev}$ is undefined at $t = 0$, $\mathit{RelTime}$ is undefined when $t + n < 0$, $\mathit{MkInterval}$ is undefined when $t_1 > t_2$, $\mathit{ValBefore}_\rho$ is undefined when no prior operation has fired, and $\mathit{Val}_\rho$ is undefined when the entity has no value at the given time point. The formalism does not yet specify how partiality propagates when one of these appears inside a larger expression or predicate. For example, if $\mathit{Now} = 0$ then $\mathit{Val}_\rho(e,\, \mathit{Prev}(\mathit{Now})) = 5$ contains an undefined sub-expression, and the truth value of the predicate is ambiguous. The options are:
    \begin{itemize}
        \item \emph{Well-formedness condition}: forbid such expressions syntactically, requiring the user to guard them explicitly (e.g.\ $\mathit{Now} > 0$).
        \item \emph{Three-valued / Kleene logic}: undefinedness propagates upward and yields a third truth value.
        \item \emph{Definedness guard}: any predicate containing an undefined sub-expression evaluates to $\mathit{false}$.
    \end{itemize}
    \item \textbf{Open question -- Channel/transmit semantics}: the transmitted value persists on the channel until the next \textit{transmit}, so \textit{receive} at any $t' \geq t$ reads the last transmitted value. It is unclear whether this is the right model for all channel types, or whether some channels should require the receiver to observe the value at the exact transmission time.
\end{itemize}

\section{Future Extensions}

\begin{itemize}
    \item \textbf{Predicate snapshot comparison}: it may be useful to capture the \emph{set of predicates that hold} at a given time point and compare that set with the one captured at a different time point. For example, one might want to assert that the set of predicates holding at time $t_1$ is equal to, a subset of, or has a non-empty intersection with the set holding at some later $t_2$. A concrete use case is a reset requirement: ``after a reset, the system must be in the same logical state as at initialisation'', where ``logical state'' is characterised by which predicates hold. This would require introducing a function $\mathit{Snapshot}_\rho : \mathcal{P}(\mathbf{Pred}) \times T \rightarrow \mathcal{P}(\mathbf{Pred})$ that returns the subset of a given predicate set that holds at a time point, and then allowing set operations ($=$, $\subseteq$, $\cap \neq \emptyset$) on the results.
    \item \textbf{Time-when expression}: add $\mathit{LastTrue}_\rho(\psi, t)$ returning the last $t' \leq t$ at which $\psi(\rho, t') = \mathit{true}$, and symmetrically $\mathit{FirstTrue}_\rho(\psi, t)$. Currently only $\mathit{LastOcc}_\rho$ and $\mathit{FirstOcc}_\rho$ cover this for \texttt{Event} entities specifically.
    \item \textbf{Time element for continuous requirements}: a requirement could optionally declare a \emph{time element} $\varepsilon \in \mathbb{R}_{>0}$, representing the smallest meaningful time increment in that requirement (e.g. a sampling period or clock cycle). In the continuous-time flavour, $\mathit{Prev}$ and $\mathit{Next}$ are currently undefined because there is no canonical predecessor or successor of a real-valued time point. With a declared $\varepsilon$, they could be lifted to continuous time as $\mathit{Prev}(t) = t - \varepsilon$ and $\mathit{Next}(t) = t + \varepsilon$, making the full set of discrete-time expressions available to requirements that have a known sampling granularity.
    \item \textbf{Unified time model via time element}: the current formalism maintains two separate time flavours, $T_D = \mathbb{N}$ and $T_C = \mathbb{R}_{\geq 0}$, selected per requirement. These could be unified into a single model by making the time element $\varepsilon$ (see the time element extension) a mandatory requirement parameter. The unified time set would then be $T = \{0,\, \varepsilon,\, 2\varepsilon,\, 3\varepsilon,\, \ldots\}$, restricting time to multiples of $\varepsilon$ and making $\mathit{Prev}(t) = t - \varepsilon$ and $\mathit{Next}(t) = t + \varepsilon$ total (except at $t = 0$). Under this model, discrete time is simply the special case $\varepsilon = 1$, and a ``continuous'' requirement with a known sampling period $\varepsilon$ is expressed the same way. Crucially, this is a stronger commitment than merely defining $\mathit{Prev}$ and $\mathit{Next}$ on an otherwise uncountable $\mathbb{R}_{\geq 0}$: it makes the time domain itself countable, so that quantifiers such as $\forall t \in T$ range over a countable set, which is important for coverage checking to remain tractable.
    \item \textbf{Open question -- Synchronous channel model}: the current model is asynchronous -- \textit{transmit} and \textit{receive} are independent operations that may occur at different times. An alternative synchronous model would have \textit{transmit} directly cause \textit{received} to fire at the same time $t$, making \textit{receive} a consequence of \textit{transmit} rather than a separate operation.
    \item \textbf{Typed sets}
    \item \textbf{Entity lookup by human-readable label.} A function $\mathit{IdByLabel} : \Gamma^* \rightharpoonup \mathit{ID}_E$ could resolve a short human-readable name to an entity identifier. The function is partial (undefined when no entity matches or the label is ambiguous). This is primarily a tooling concern: the formal handle is the entity identifier $id_e$; how a user locates that identifier from natural language is outside the core formalism.
    \item \textbf{Required modifier keys.} The current $M_E$ function specifies which modifier keys are \emph{permitted} for each entity type, but all modifiers are optional. A future extension could introduce a companion function $M_E^{\text{req}} : T_E \rightarrow \mathcal{P}(\mathit{ModKey})$ specifying which modifier keys are \emph{required}. For example, predefined-event \texttt{Event} entities (those used with $\mathit{EvtPred}$) should always have both $\mathit{target}$ and $\mathit{type}$ defined.
    \item \textbf{Aggregate expression recognition.} When a test case computes a result through a sequence of arithmetic operations (e.g.\ summing values and dividing by a count), the coverage checker must be able to recognise that the computation is equivalent to an aggregate such as average or median. Without this, a test exercising $\mathit{avg}(e, i)$ expressed via raw arithmetic would not be matched against a requirement that references the same aggregate by name. A future extension should define canonical aggregate expressions ($\mathit{Avg}$, $\mathit{Median}$, etc.) and a normalisation or equivalence procedure that maps common arithmetic patterns to them.
    \item \textbf{Dynamic entity lifetimes.} The entity set $E$ is fixed for the scope of a requirement, so the formalism cannot distinguish between an entity that does not exist at time $t$ and one that merely has no value there -- both reduce to $\sigma(e)$ being undefined. If entities with dynamic lifetimes are needed (e.g.\ a spawned process or an opened connection), an explicit existence predicate $\mathit{Exists} : E \times T \to \mathbb{B}$ should be introduced, with $\mathit{Val}_\rho(e, t)$ defined only when $\mathit{Exists}(e, t) = \mathit{true}$.
    \item \textbf{User-defined modifier keys.} Currently, the set of permitted modifier keys for each entity type is fixed by $M_E$ and can only be extended at the formalism level. A future extension could allow users to attach arbitrary key-value annotations to entities inline when writing a requirement -- for example, to record a physical unit, an address, or a domain-specific tag -- without requiring a change to $M_E$. One approach is to introduce a distinguished \emph{user-defined} modifier namespace $\mathit{ModKey}_{\mathit{user}} \cap \mathit{ModKey} = \emptyset$ and extend $M_E$ so that every entity type implicitly permits all keys from $\mathit{ModKey}_{\mathit{user}}$. The \texttt{Abstract} entity type already serves as a catch-all for entities that do not fit a predefined kind; user-defined modifiers would naturally complement it.
\end{itemize}

\newpage

\input{examples}

\end{document}