% FORMALISM OVERVIEW
%
% PURPOSE: Define a logic-agnostic formal language for expressing embedded/avionics
% software requirements and test cases, so that coverage gaps can be detected
% mechanically. The ten Embraer requirements in req/ are the concrete test bed.
%
% AUTHORITATIVE REFERENCE: ../abstract-final.md (8-section spec).
% This file is the LaTeX rendering of that spec.
%
% KEY DESIGN PRINCIPLES:
%
% 1. LAYERS
%    Primitives (T, V) -> Entities & Dictionary -> Valuations & Traces ->
%    Expressions & Predicates -> Operations -> Temporal Constructs -> Requirements
%    -> Test Cases & Coverage
%
% 2. EXPRESSIONS ARE EVALUATED AT AMBIENT (rho, t)
%    Every expression is implicitly a function of trace rho and current time t.
%    e.g. prev(e) means Val_rho(e, t-1) where t is supplied by the surrounding
%    temporal operator, NOT by the operation call. This means operations do NOT
%    take t as an explicit argument -- t is implicit.
%
% 3. OPERATIONS DEFINE POINT-IN-TIME EFFECT PREDICATES
%    An operation such as calculate(e, expr) produces a predicate that holds at
%    the *current* t. The temporal extent (always, eventually, from t onwards)
%    is specified at the requirement level via temporal operators (always, triggered,
%    etc.), not inside the operation definition.
%    Example: always(calculate(e, 1.2 * prev(e))) expresses the recurrence
%    e(k) = 1.2 * e(k-1) for all k -- the always handles "for all time".
%
% 4. THE DICTIONARY IS OPEN
%    Entity types (Signal, Storage, Event, Channel, State, Value, Mapping, Abstract)
%    each declare: semantics, applicable operations, and admitted modifiers.
%    New types are added by extending the dictionary, not by changing the core.
%
% 5. COVERAGE IS DERIVED FROM REQUIREMENT STRUCTURE
%    MC/DC for ALL_OF conditions, per-element for ANY_OF, boundary tests for time
%    bounds, negative cases for guards, ordering violation tests for strict sequences.
%
% KNOWN OPEN ISSUES (see Notes section and todo.md):
%    - Discrete vs continuous time flavour handling
%    - Partiality of time functions (Prev at t=0)

% Requirements for this docuement:
% - it needs to be written in simple wording, dropping any fluff

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{stmaryrd}
\usepackage[margin=2.5cm]{geometry}
\usepackage[normalem]{ulem}
\usepackage{nameref}
\usepackage{etoolbox}

\newtheorem{definition}{Definition}
\AtEndEnvironment{definition}{\vspace{0.5em}}

\title{PP1 - Evaluating Test Coverage Against Software Requirements -- formal base}
\author{  }
\date{April 2026}

\begin{document}

\maketitle

\section{Introduction}

\subsection{The purpose of the formal base}

This document defines a logic-agnostic formal base for expressing software requirements and test cases.
The goal is to enable mechanical detection of coverage gaps: given a set of requirements and a set of test cases, the checker determines which requirement conditions are not exercised by any test.

\subsection{The design principles of the formal base}

The formalism is designed around two mandatory properties.

\begin{itemize}
\item \textbf{extensibility} -- adding new entity kinds or operations must require only extending the entity type set $T_E$, the values set $V$, the entity type modifiers function $M_E$, and the entity type operations function $O_T$.
\item \textbf{matchability} -- a formalised requirement and a formalised test case must be expressed using the same \textbf{predicates}, so that the coverage checker can determine which truth assignments to those predicates are exercised by the tests and which logical combinations remain uncovered.
\end{itemize}

\section{The primitives}

These primitive sets serve as the base that the rest of the formalism builds on. The primitive sets are the following:
\begin{itemize}
    \item $T_D$, $T_C$ -- the \textit{discrete time} and \textit{continuous time} sets.
          An infinite totally ordered set; the time set can have \textit{discrete} or \textit{continuous} flavour; exactly one flavour is chosen per requirement.
    \item $V$ -- the $Values$ set. These can be used to describe the values of entities.
\end{itemize}

\begin{definition}[Discrete time set]\label{def:time_discrete}
    The \emph{discrete time set} $T_D$ is totally ordered and countable.
    \[
        T_D = \mathbb{N}
    \]
\end{definition}

\begin{definition}[Continuous time set]\label{def:time_continuous}
    The \emph{continuous time set} $T_C$ is totally ordered and uncountable.
    \[
        T_C = \mathbb{R}_{\geq 0}
    \]
\end{definition}

\begin{definition}[Duration]\label{def:duration}
    A \emph{duration} is a non-negative element of the time set, representing the length of a time span:
    \[
        \Delta = T
    \]
    Thus $\Delta = \mathbb{N}$ for discrete time and $\Delta = \mathbb{R}_{\geq 0}$ for continuous time.
\end{definition}

\begin{definition}[Interval]\label{def:interval}
    An \emph{interval} is a pair of a start time point and a duration:
    \[
        \mathcal{I} = T \times \Delta
    \]
    An interval $(t, d) \in \mathcal{I}$ represents the time span $[t,\, t + d]$.
\end{definition}

\begin{definition}[Values]\label{def:values}
    The \emph{values set} $V$ is the union of the following sets:
    \[
        V = \mathbb{B} \cup \mathbb{R} \cup T \cup \mathcal{I}
    \]
    where $\mathbb{B} = \{\mathit{true}, \mathit{false}\}$ is the set of booleans, $\mathbb{R}$ is the set of real numbers, $T$ is the time set, and $\mathcal{I}$ is the set of intervals.
\end{definition}

\section{Entities and entity types}

An entity type is an item belonging to a set of predefined entity types.

\begin{definition}[Entity type set]\label{def:entity_type_set}
    \[
        T_\text{E} = \{ \text{Signal}, \text{Storage}, \text{Event}, \text{Channel}, \text{State}, \text{Value}, \text{Abstract}\}
    \]
\end{definition}

\begin{definition}[Entity name set]\label{def:entity_name_set}
    \[
        N_\text{E} \subseteq \Sigma^*
    \]

    , where $\Sigma$ is an alphabet.
\end{definition}

Each entity type has a set of \emph{predefined events} — events that fire automatically as a result of operations on the entity, without requiring explicit user declaration.

\begin{definition}[Predefined events function]\label{def:predefined_events}
    $PE : T_E \rightarrow \mathcal{P}(\mathit{EventKind})$ maps each entity type to its set of predefined event kinds.
    A predefined event instance for entity $e$ is written $ek(e)$, where $ek \in PE(t_e)$.
    The predefined event kinds are:
    \begin{gather*}
        PE(\text{Signal}) = \{\, \mathit{calculated} \,\} \qquad
        PE(\text{Storage}) = \{\, \mathit{written},\, \mathit{read} \,\} \\
        PE(\text{State}) = \{\, \mathit{entered} \,\} \qquad
        PE(\text{Channel}) = \{\, \mathit{transmitted},\, \mathit{received} \,\}
    \end{gather*}
    All other entity types have no predefined events.
\end{definition}

Entity type modifiers are properties which can be attached to an entity of a particular type.
A modifier is either a \emph{flag} (whose presence alone carries meaning) or a \emph{parameterised modifier} (which additionally carries a value in $V$).

Each entity type is compatible with only a subset of the modifier keys.
Therefore we define the \textit{entity type modifiers function}.

\begin{definition}[Entity type modifiers function]\label{def:entity_type_modifiers_fn}
    \[
        M_\text{E}: T_E \rightarrow \mathcal{P}(\mathit{ModKey})
    \]

    , where $\mathit{ModKey}$ is the set of modifier keys and $\mathcal{P}(\mathit{ModKey})$ is the power set of $\mathit{ModKey}$.
\end{definition}

An entity is something that is referenced in the requirement.
Formally, entity is defined as a four-item tuple.

\begin{definition}[Entity]\label{def:entity}
    \[
        e = (n_e, t_e, \mu_e, d_e)
    \]

    , where
    \begin{itemize}
        \item $n_e$ is the name of the entity; $n_e \in N_E$
        \item $t_e$ is the type of the entity; $t_e \in T_E$
        \item $\mu_e$ is the modifier assignment function of the entity; $\mu_e : \mathit{ModKey} \rightharpoonup V$, $\mathrm{dom}(\mu_e) \subseteq M_E(t_e)$ \footnote{$\mathrm{dom}(\mu_e)$ is a subset of the set of modifier keys compatible with the entity type $t_e$. It is not required that a value is defined for each modifier key compatible with $t_e$.}
        \item $d_e$ is the description of the entity; $d_e \in \Sigma^*$
    \end{itemize}, where $dom(\mu_e)$ denotes the domain of the function $\mu_e$. We write $E$ for the set of all entities. The set $E$ is finite.\footnote{The set $E$ is finite in the scope of a single requirement.}
\end{definition}



\section{Valuations and traces}

\begin{definition}[Valuation]\label{def:valuation}
    A valuation is a partial function $\sigma : E \rightharpoonup V$ assigning a value to a subset of entities at each time point\footnote{For instance a register entity can have its value undefined at time points preceeding a write to this register.}; $\sigma(e)$ is undefined when the entity has no defined value at that time point.
    For entities of type \texttt{Event}, $\sigma(e)$ is always defined and $\sigma(e) \in \{\mathit{true}, \mathit{false}\}$,
    where $\sigma(e) = \mathit{true}$ iff the event occurs at that time point.
\end{definition}

We write $\Sigma$ for the set of all valuations.

\begin{definition}[Trace]\label{def:trace}
    A trace is a function $\rho : T \to \Sigma$. We write $\rho(t)(e)$ for the value of entity $e$ at time $t$.
\end{definition}

We write $\Sigma^T$ for the set of all traces, i.e.\ the set of all functions from $T$ to $\Sigma$.

\footnote{This is standard function space notation: $\Sigma^T = \{\, \rho \mid \rho : T \to \Sigma \,\}$.}

\section{Expressions and predicates}

All expression functions defined in this section evaluate to a value in $V$.
We write $\mathbf{Expr}$ for the set of all expressions.
Functions in \S\ref{subsec:time} (\nameref{subsec:time}) depend only on time points or intervals.
Functions in \S\ref{subsec:trace} (\nameref{subsec:trace}) additionally take a trace $\rho \in \Sigma^T$
and produce values that reflect the state of the system over time.

\subsection{Expressions}\label{subsec:expressions}

\subsubsection{Constants and arithmetic}\label{subsec:constants}

\begin{definition}[Constant]\label{def:constant}
    Any value $c \in V$ may appear as a constant expression.
\end{definition}

The standard arithmetic operators $+$, $-$, $\times$, $/$ are defined on numeric values ($\mathbb{R}$ and its subsets) in the usual sense.

\subsubsection{Time functions}\label{subsec:time}

The following functions operate on time points and intervals only; they do not depend on a trace.

\begin{definition}[Current time]\label{def:now}
    $\mathit{Now} \in T$ returns the ambient time point $t$:
    \[
        \mathit{Now} \;=\; t
    \]
\end{definition}

The following two functions are defined for \textbf{discrete time only} ($T = T_D = \mathbb{N}$).

\begin{definition}[]\label{def:prev}
    $\mathit{Prev} : T_D \rightharpoonup T_D$ returns the preceding time step:
    \[
        \mathit{Prev}(t) \;=\; t - 1
    \]
    The function is partial: it is undefined at $t = 0$ since $t - 1 \notin T_D$.
\end{definition}

\begin{definition}[]\label{def:next}
    $\mathit{Next} : T_D \rightarrow T_D$ returns the following time step:
    \[
        \mathit{Next}(t) \;=\; t + 1
    \]
\end{definition}

\begin{definition}[]\label{def:reltime}
    $\mathit{RelTime} : T_D \times \mathbb{Z} \rightharpoonup T_D$ shifts a time point by $n$ steps, where $n \in \mathbb{Z}$ may be positive or negative:
    \[
        \mathit{RelTime}(t,\, n) \;=\; t + n
    \]
    The function is partial: it is undefined when $t + n < 0$ since $t + n \notin T_D$.
    Note that $\mathit{Prev}(t) = \mathit{RelTime}(t, -1)$ and $\mathit{Next}(t) = \mathit{RelTime}(t, 1)$.
\end{definition}

The following functions are defined for both time flavours.

\begin{definition}[]\label{def:diff}
    $\mathit{Diff} : T \times T \rightarrow \Delta$ returns the duration between two time points:
    \[
        \mathit{Diff}(t_1,\, t_2) \;=\; |t_1 - t_2|
    \]
\end{definition}

\begin{definition}[]\label{def:mk_interval}
    $\mathit{MkInterval} : T \times T \rightharpoonup \mathcal{I}$ constructs an interval from a start and end time point:
    \[
        \mathit{MkInterval}(t_1,\, t_2) \;=\; (t_1,\; t_2 - t_1)
    \]
    The function is partial: it is undefined when $t_1 > t_2$.
\end{definition}

\begin{definition}[Since-zero interval]\label{def:since_zero}
    $\mathit{SinceZero} : T \rightarrow \mathcal{I}$ constructs an interval from time zero to a given time point $t$:
    \[
        \mathit{SinceZero}(t) \;=\; (0,\; t)
    \]
\end{definition}

\begin{definition}[Interval start]\label{def:interval_start}
    $\mathit{Start} : \mathcal{I} \rightarrow T$ returns the start time point of an interval:
    \[
        \mathit{Start}((t,\, d)) \;=\; t
    \]
\end{definition}

\begin{definition}[Interval end]\label{def:interval_end}
    $\mathit{End} : \mathcal{I} \rightarrow T$ returns the end time point of an interval:
    \[
        \mathit{End}((t,\, d)) \;=\; t + d
    \]
\end{definition}

\begin{definition}[Interval duration]\label{def:interval_duration}
    $\mathit{Duration} : \mathcal{I} \rightarrow \Delta$ returns the duration of an interval:
    \[
        \mathit{Duration}((t,\, d)) \;=\; d
    \]
\end{definition}

\subsubsection{Entity attribute accessors}\label{subsec:entity_attrs}

These expressions return static properties of entities and do not depend on the trace or time.

% TODO: this needs to be fixed

\begin{definition}[Address]\label{def:addr}
    For a \texttt{Storage} entity $e \in E$ carrying an address modifier, $\mathit{Addr} : E \rightarrow V$ returns the static address value:
    \[
        \mathit{Addr}(e) \;=\; \mu_e(\mathit{address})
    \]
\end{definition}

\subsubsection{Trace functions}\label{subsec:trace}

Expressions are terms that evaluate to a value in $V$ given a trace $\rho$.

\begin{definition}[]\label{def:expr_val}
    For a trace $\rho \in \Sigma^T$, the partial function $\mathit{Val}_\rho: E \times T \rightharpoonup V$ denotes the value of entity $e \in E$ at time $t \in T$:
    \[
        \mathit{Val}_\rho(e,\, t) \;=\; \rho(t)(e)
    \]
    The function is partial: $\mathit{Val}_\rho(e,\, t)$ is undefined when $e \notin \mathrm{dom}(\rho(t))$.
\end{definition}

\begin{definition}[Event occurrence count]\label{def:evt_occ_count}
    For a trace $\rho \in \Sigma^T$ and an entity $ev \in E$ of type \texttt{Event},
    $\mathit{EvtOccCount}_\rho : E \times \mathcal{I} \rightarrow \mathbb{N}$ returns the number of occurrences of $ev$ within an interval $i \in \mathcal{I}$:
    \[
        \mathit{EvtOccCount}_\rho(ev,\, i) \;=\; \bigl|\{\, s \in T \mid \mathit{Start}(i) \leq s \leq \mathit{End}(i) \;\land\; \mathit{Val}_\rho(ev,\, s) = \mathit{true} \,\}\bigr|
    \]
\end{definition}

\begin{definition}[Last occurrences function]\label{def:last_occ}
    For a trace $\rho \in \Sigma^T$, an entity $ev \in E$ of type \texttt{Event}, a time point $t \in T$, and $n \in \mathbb{N}^+$,
    let $t_1 \leq t_2 \leq \cdots \leq t_n$ be the $n$ most recent occurrence times of $ev$ up to $t$, i.e.\ the $n$ largest elements of $\{\, s \in T \mid s \leq t \;\land\; \mathit{Val}_\rho(ev,\, s) = \mathit{true} \,\}$ sorted in ascending order.
    Then $\mathit{LastOcc}_\rho : E \times T \times \mathbb{N}^+ \rightharpoonup \mathcal{I}$ is defined as:
    \[
        \mathit{LastOcc}_\rho(ev,\, t,\, n) \;=\; \mathit{MkInterval}(t_1,\, t_n)
    \]
    The function is partial: it is undefined when fewer than $n$ occurrences of $ev$ exist up to $t$.

    \textit{Note:} The time of a single occurrence is obtained via $\mathit{LastOcc}_\rho(ev, t, 1)$, which returns an interval of duration $0$. The occurrence time is then $\mathit{Start}(\mathit{LastOcc}_\rho(ev, t, 1))$.
\end{definition}

\begin{definition}[First occurrences function]\label{def:first_occ}
    For a trace $\rho \in \Sigma^T$, an entity $ev \in E$ of type \texttt{Event}, a time point $t \in T$, and $n \in \mathbb{N}^+$,
    let $t_1 \leq t_2 \leq \cdots \leq t_n$ be the $n$ earliest occurrence times of $ev$ up to $t$, i.e.\ the $n$ smallest elements of $\{\, s \in T \mid s \leq t \;\land\; \mathit{Val}_\rho(ev,\, s) = \mathit{true} \,\}$ sorted in ascending order.
    Then $\mathit{FirstOcc}_\rho : E \times T \times \mathbb{N}^+ \rightharpoonup \mathcal{I}$ is defined as:
    \[
        \mathit{FirstOcc}_\rho(ev,\, t,\, n) \;=\; \mathit{MkInterval}(t_1,\, t_n)
    \]
    The function is partial: it is undefined when fewer than $n$ occurrences of $ev$ exist up to $t$.

    \textit{Note:} The time of the first occurrence is $\mathit{Start}(\mathit{FirstOcc}_\rho(ev, t, 1))$.
\end{definition}

\begin{definition}[Maximum value function]\label{def:max_val}
    For a trace $\rho \in \Sigma^T$ and an entity $e \in E$ with an ordered value domain,
    $\mathit{MaxVal}_\rho : E \times \mathcal{I} \rightarrow V$ returns the maximum value of $e$ over an interval $i \in \mathcal{I}$:
    \[
        \mathit{MaxVal}_\rho(e,\, i) \;=\; \sup\{\, \mathit{Val}_\rho(e,\, s) \mid s \in T,\; \mathit{Start}(i) \leq s \leq \mathit{End}(i) \,\}
    \]
\end{definition}

\begin{definition}[Minimum value function]\label{def:min_val}
    For a trace $\rho \in \Sigma^T$ and an entity $e \in E$ with an ordered value domain,
    $\mathit{MinVal}_\rho : E \times \mathcal{I} \rightarrow V$ returns the minimum value of $e$ over an interval $i \in \mathcal{I}$:
    \[
        \mathit{MinVal}_\rho(e,\, i) \;=\; \inf\{\, \mathit{Val}_\rho(e,\, s) \mid s \in T,\; \mathit{Start}(i) \leq s \leq \mathit{End}(i) \,\}
    \]
\end{definition}

\subsection{Predicates}\label{subsec:predicates}

A predicate is a function that evaluates to a value in $\mathbb{B}$.
We write $\mathbf{Pred}$ for the set of all predicates.

\begin{definition}[Boolean constants]\label{def:pred_bool}
    $\top \in \mathbf{Pred}$ and $\bot \in \mathbf{Pred}$ are the always-true and always-false predicates respectively.
\end{definition}

\begin{definition}[Comparison predicate]\label{def:pred_cmp}
    For expressions $a, b \in V$ and a comparison operator $\mathit{op} \in \{=,\, \neq,\, <,\, \leq,\, >,\, \geq\}$:
    \[
        \mathit{Cmp}(a,\, \mathit{op},\, b) \;=\; a \;\mathit{op}\; b
    \]
    The operators $<, \leq, >, \geq$ require $a, b$ to be from an ordered domain.
    Since $\Delta \subseteq \mathbb{R}$, durations are an ordered domain and duration comparison is a valid instance of $\mathit{Cmp}$.
\end{definition}

\begin{definition}[Negation]\label{def:pred_not}
    For a predicate $P \in \mathbf{Pred}$:
    \[
        \mathit{Not}(P) \;=\; \lnot P
    \]
\end{definition}

\begin{definition}[Conjunction and disjunction]\label{def:pred_allandany}
    For a finite set of predicates $C \subseteq \mathbf{Pred}$:
    \[
        \mathit{AllOf}(C) \;=\; \bigwedge_{P \in C} P \qquad
        \mathit{AnyOf}(C) \;=\; \bigvee_{P \in C} P
    \]
\end{definition}

\begin{definition}[Implication]\label{def:pred_implies}
    For predicates $A, B \in \mathbf{Pred}$:
    \[
        \mathit{Implies}(A,\, B) \;=\; A \Rightarrow B
    \]
    where $\Rightarrow$ is the standard logical implication.
\end{definition}

\begin{definition}[Quantifiers]\label{def:pred_quant}
    For a finite set $S$ and a predicate $P(x)$ parameterised by $x \in S$:
    \[
        \mathit{ForAll}(S,\, P) \;=\; \bigwedge_{x \in S} P(x) \qquad
        \mathit{Exists}(S,\, P) \;=\; \bigvee_{x \in S} P(x)
    \]
\end{definition}

\begin{definition}[Event happening]\label{def:pred_happening}
    For an entity $e \in E$ with $t_e = \text{Event}$ and a time point $t \in T$:
    \[
        \mathit{Happening}(e,\, t) \;\Leftrightarrow\; \mathit{Val}_\rho(e,\, t) = \mathit{true}
    \]
\end{definition}

\begin{definition}[Event has happened]\label{def:pred_has_happened}
    For an entity $e \in E$ with $t_e = \text{Event}$ and a time point $t \in T$:
    \[
        \mathit{HasHappened}_\rho(e,\, t) \;\Leftrightarrow\; \exists t' \leq t:\; \mathit{Happening}(e,\, t')
    \]
\end{definition}

The following predicates operate on intervals and durations.
Since $\Delta \subseteq \mathbb{R}$, duration comparison is a special case of $\mathit{Cmp}$ (Definition~\ref{def:pred_cmp}).

\begin{definition}[Interval inclusion]\label{def:pred_includes}
    Interval $i_1$ includes $i_2$ iff $i_2$ is entirely contained within $i_1$:
    \[
        \mathit{IntervalIncludes}(i_1,\, i_2) \;\Leftrightarrow\;
        \mathit{Start}(i_1) \leq \mathit{Start}(i_2) \;\land\; \mathit{End}(i_2) \leq \mathit{End}(i_1)
    \]
\end{definition}

\begin{definition}[Interval exclusion]\label{def:pred_excludes}
    Intervals $i_1$ and $i_2$ are exclusive iff they are completely disjoint:
    \[
        \mathit{IntervalExcludes}(i_1,\, i_2) \;\Leftrightarrow\;
        \mathit{End}(i_1) < \mathit{Start}(i_2) \;\lor\; \mathit{End}(i_2) < \mathit{Start}(i_1)
    \]
\end{definition}

\begin{definition}[Start included]\label{def:pred_start_in}
    The start of $i_1$ falls within $i_2$:
    \[
        \mathit{StartOfFirstIntervalIn}(i_1,\, i_2) \;\Leftrightarrow\;
        \mathit{Start}(i_2) \leq \mathit{Start}(i_1) \leq \mathit{End}(i_2)
    \]
\end{definition}

\begin{definition}[End included]\label{def:pred_end_in}
    The end of $i_1$ falls within $i_2$:
    \[
        \mathit{EndOfFirstIntervalIn}(i_1,\, i_2) \;\Leftrightarrow\;
        \mathit{Start}(i_2) \leq \mathit{End}(i_1) \leq \mathit{End}(i_2)
    \]
\end{definition}

% DESIGN CONCLUSION: OPERATIONS AND TEMPORAL EXTENT
%
% Operations (e.g. calculate, fire) do NOT take t as an explicit argument.
% t is "ambient" -- it is supplied by the surrounding temporal construct
% (Always, Triggered, etc.) at the requirement level.
%
% An operation produces a point-in-time effect predicate that holds at the
% ambient t only. It says nothing about any other time point.
%
% We considered making calculate's effect hold from t onward
% (i.e. forall t' >= t: Val_rho(e, t') = expr), but rejected this because:
%   (a) The trace rho is a fixed mathematical object -- operations do not
%       "write" to it. It already has whatever values it has at every time point.
%   (b) A later operation in the same requirement may set a different value,
%       which would directly contradict a "forall t' >= t" clause.
%
% Temporal extent is always the responsibility of the surrounding temporal
% construct:
%   Always(calculate(e, expr))           -- holds at every t
%   Causes(cond, calculate(e, expr))     -- holds at every t where cond fires
%   forall t' >= t: ...                  -- expressed at requirement level, not
%                                           inside the operation definition

\section{Operations}

An operation is an action that the system performs at a specific point in time.
Each operation has a set of arguments and produces an \emph{effect predicate} — a formal statement, parameterised by the firing time $t$, of what must hold in the trace as a result of the operation being performed.
The operation is applied to some entity, which means that the argument list of every operation must include the entity the operation is applied to.

\begin{definition}[Operation]\label{def:operation}
    An \emph{operation} is a function
    \[
        op : T \times E \times A_1 \times \cdots \times A_n \rightarrow \mathbf{Pred}
    \]
    for some $n \geq 0$ and additional argument types $A_i \subseteq V \cup E$, where $T$ is the time set and $E$ is the mandatory target entity.
    The predicate $op(t,\, e,\, a_1,\, \ldots,\, a_n)$ is the \emph{effect} of $op$ applied to entity $e$ at time $t$ -- it describes the condition that the operation causes to hold.
    The trigger condition that invokes an operation is specified at the requirement level.
    We write $O$ for the set of all operations.
\end{definition}

\begin{definition}[Entity type operations function]\label{def:entity_type_ops_fn}
    \[
        O_T : T_E \times \mathcal{P}(M) \rightarrow \mathcal{P}(O)
    \]
    maps each entity type and set of modifiers to its applicable operations.\footnote{$O = \bigcup_{\tau \in T_E,\, \mu \subseteq M} O_T(\tau,\, \mu)$.}

    The concrete mapping, including argument types and effects, is as follows.
    In effect predicates, $t_{\mathit{next}}$ denotes the next time the same operation fires on the same entity after $t$, or $\infty$ if no such firing occurs.

    \medskip
    \renewcommand{\arraystretch}{2}
    \begin{tabularx}{\textwidth}{lXX}
        \toprule
        \textbf{Operation} & \textbf{Notes}                                                                    & \textbf{Effect predicates} \\
        \midrule
        $\mathit{calculate}(t,\, e,\, \mathit{expr})$
                           & $t \in T$,\newline $e \in E,\ t_e = \text{Signal}$,\newline $\mathit{expr} \in \mathbf{Expr}$
                           & $\mathit{Happening}(\mathit{calculated}(e),\, t)$ \newline $\forall t' \in [t,\, t_{\mathit{next}}):\; \mathit{Val}_\rho(e,\, t') = \mathit{expr}(\rho,\, t)$                  \\[4pt]
        $\mathit{write}(t,\, e,\, \mathit{expr})$
                           & $t \in T$,\newline $e \in E,\ t_e = \text{Storage}$,\newline $\mathit{expr} \in \mathbf{Expr}$
                           & $\mathit{Happening}(\mathit{written}(e),\, t)$ \newline $\forall t' \in [t,\, t_{\mathit{next}}):\; \mathit{Val}_\rho(e,\, t') = \mathit{expr}(\rho,\, t)$                  \\[4pt]
        $\mathit{read}(t,\, e_{\mathit{src}},\, e_{\mathit{dst}})$
                           & $t \in T$,\newline $e_{\mathit{src}} \in E,\ t_{e_{\mathit{src}}} = \text{Storage}$,\newline $e_{\mathit{dst}} \in E,\ t_{e_{\mathit{dst}}} = \text{Signal}$
                           & $\mathit{Happening}(\mathit{read}(e_{\mathit{src}}),\, t)$ \newline $\forall t' \in [t,\, t_{\mathit{next}}):\; \mathit{Val}_\rho(e_{\mathit{dst}},\, t') = \mathit{Val}_\rho(e_{\mathit{src}},\, t)$  \\[4pt]
        $\mathit{fire}(t,\, e)$
                           & $t \in T$, $e \in E,\ t_e = \text{Event}$
                           & $\mathit{Happening}(e,\, t)$                                                                                            \\[4pt]
        $\mathit{set\_state}(t,\, e,\, s)$
                           & $t \in T$,\newline $e \in E,\ t_e = \text{State}$,\newline $s \in V$
                           & $\mathit{Happening}(\mathit{entered}(e),\, t)$ \newline $\forall t' \in [t,\, t_{\mathit{next}}):\; \mathit{Val}_\rho(e,\, t') = s$  \\[4pt]
        $\mathit{transmit}(t,\, e,\, \mathit{expr})$
                           & $t \in T$,\newline $e \in E,\ t_e = \text{Channel}$,\newline $\mathit{expr} \in \mathbf{Expr}$
                           & $\mathit{Happening}(\mathit{transmitted}(e),\, t)$ \newline $\forall t' \in [t,\, t_{\mathit{next}}):\; \mathit{Val}_\rho(e,\, t') = \mathit{expr}(\rho,\, t)$      \\[4pt]
        $\mathit{receive}(t,\, e_{\mathit{src}},\, e_{\mathit{dst}})$
                           & $t \in T$,\newline $e_{\mathit{src}} \in E,\ t_{e_{\mathit{src}}} = \text{Channel}$,\newline $e_{\mathit{dst}} \in E,\ t_{e_{\mathit{dst}}} = \text{Signal}$
                           & $\mathit{Happening}(\mathit{received}(e_{\mathit{src}}),\, t)$ \newline $\mathit{Val}_\rho(e_{\mathit{dst}},\, t) = \mathit{Val}_\rho(e_{\mathit{src}},\, t)$  \\
        \bottomrule
    \end{tabularx}
    \medskip

    % \noindent Here $\mathit{ValueAt}(a)$ denotes the value stored at memory address $a$.
\end{definition}

\section{Temporal constructs}

So far, predicates ($\psi \in \mathbf{Pred}$) are point-in-time: they say whether something holds at a specific time $t$ in a specific trace $\rho$.
A \emph{trace predicate} lifts this to the whole trace: it says whether something holds across the entire execution, not just at one moment.
Formally, a trace predicate is a function $P : \Sigma^T \to \mathbb{B}$ — it takes a complete trace and returns true or false.
A trace $\rho$ \emph{satisfies} a requirement $R$ (written $\rho \models R$) iff $R$'s constraint, expressed as a trace predicate $P$, gives $P(\rho) = \mathit{true}$.

The temporal constructs below build trace predicates from point-in-time predicates $\psi \in \mathbf{Pred}$ by quantifying over time — expressing things like ``$\psi$ must hold at all times'' or ``whenever $\psi_{\mathit{cond}}$ holds, $\psi_{\mathit{effect}}$ must eventually follow.''

\begin{definition}[Trace predicate]\label{def:trace_predicate}
    A \emph{trace predicate} is a function $P : \Sigma^T \to \mathbb{B}$.
    We write $\mathbf{TracePred}$ for the set of all trace predicates.
    A trace $\rho$ \emph{satisfies} a trace predicate $P$ (written $\rho \models P$) iff $P(\rho) = \mathit{true}$.
\end{definition}

\begin{definition}[Temporal constructs]\label{def:temporal_constructs}
    \begin{align*}
        \mathit{Always}(\psi) \;&\Leftrightarrow\; \forall t \in T:\; \psi(\rho, t) \\
        \mathit{Eventually}(\psi) \;&\Leftrightarrow\; \exists t \in T:\; \psi(\rho, t) \\
        \mathit{Initial}(\psi) \;&\Leftrightarrow\; \psi(\rho, 0) \\
        \mathit{Causes}(\psi_{\mathit{cond}},\, \psi_{\mathit{effect}}) \;&\Leftrightarrow\;
            \forall t \in T:\; \psi_{\mathit{cond}}(\rho, t) \Rightarrow \exists t' \geq t:\; \psi_{\mathit{effect}}(\rho, t') \\
        \mathit{CausesWithin}(\psi_{\mathit{cond}},\, \psi_{\mathit{effect}},\, d) \;&\Leftrightarrow\;
            \forall t \in T:\; \psi_{\mathit{cond}}(\rho, t) \Rightarrow \exists t' \in [t,\, t+d]:\; \psi_{\mathit{effect}}(\rho, t') \\
        \mathit{Sequence}(\psi_1,\, \ldots,\, \psi_n) \;&\Leftrightarrow\;
            \exists t_1 \leq t_2 \leq \cdots \leq t_n \in T:\; \psi_i(\rho,\, t_i) \text{ for all } i \\
        \mathit{Precedes}(\psi_1,\, \psi_2) \;&\Leftrightarrow\;
            \forall t_1 \in T:\; \psi_2(\rho,\, t_1) \Rightarrow \exists t_2 \leq t_1:\; \psi_1(\rho,\, t_2) \\
        \mathit{Excludes}(\psi_1,\, \psi_2) \;&\Leftrightarrow\;
            \forall t \in T:\; \lnot\bigl(\psi_1(\rho,\, t) \land \psi_2(\rho,\, t)\bigr) \\
        \mathit{Immediately}(\psi_1,\, \psi_2) \;&\Leftrightarrow\;
            \forall t_1 \in T:\; \psi_1(\rho,\, t_1) \Rightarrow \psi_2(\rho,\, \mathit{Next}(t_1))
    \end{align*}
    where $\psi,\, \psi_{\mathit{cond}},\, \psi_{\mathit{effect}},\, \psi_1, \ldots, \psi_n \in \mathbf{Pred}$ are point-in-time predicates and $d \in \Delta$ is a duration bound.
\end{definition}

The constructs compose via standard logical operators on trace predicates:
conjunction $P_1 \land P_2$, disjunction $P_1 \lor P_2$, negation $\lnot P$,
implication $P_1 \Rightarrow P_2$ (if $P_1$ holds for trace $\rho$ then $P_2$ must also hold), and
biconditional $P_1 \Leftrightarrow P_2$.
These operate on whole traces, not at individual time points.

\section{Requirements \& Test cases}

Both requirements and test cases operate over the same shared entity set $E$, which is what makes coverage checking possible.

A requirement is a formal constraint on the system's behaviour, expressed as a trace predicate.

\begin{definition}[Requirement]\label{def:requirement}
    A \emph{requirement} is a triple $R = (\mathit{flavour},\, \mathit{entities},\, \mathit{constraint})$ where:
    \begin{itemize}
        \item $\mathit{flavour} \in \{\text{Discrete},\, \text{Continuous}\}$ fixes the time set $T$ for this requirement ($T_D$ or $T_C$ respectively), determining which expressions and temporal constructs are well-typed
        \item $\mathit{entities} \subseteq E$ is the set of entities participating in this requirement
        \item $\mathit{constraint} \in \mathbf{TracePred}$ is the formal constraint on the trace
    \end{itemize}
    A trace $\rho$ \emph{satisfies} $R$ (written $\rho \models R$) iff $\mathit{constraint}(\rho) = \mathit{true}$.
\end{definition}

A test case exercises a subset of the same entities to produce a concrete trace, which the coverage checker then analyses against the requirement's constraint.

\begin{definition}[Test case]\label{def:test_case}
    A \emph{test case} is a pair $\mathit{TestCase} = (\mathit{setup},\, \mathit{stimuli})$ where:
    \begin{itemize}
        \item $\mathit{setup}$ is a valuation $\sigma_0 : E \rightharpoonup V$ establishing the initial state of the relevant entities
        \item $\mathit{stimuli} : T \to \mathcal{P}(O)$ maps each time point to the set of operations that fire at that time; it has finite support (non-empty for only finitely many $t$)
    \end{itemize}
\end{definition}

Together, $\mathit{setup}$ and $\mathit{stimuli}$ define a trace $\rho$. The coverage checker determines which predicates in a requirement's constraint are exercised by $\rho$.

\section{Notes}

\begin{itemize}
    \item Lepsi ponechat typ entity, protoze musime mapovat na modifiers,
          item ktere jsou rozdilne pro kazdy typ entity. Zaroven muzeme namapovat
          typ entity na mnozinu operaci.
    \item Modifiers need to be parameterized.
    \item We could theoretically map entity types directly to predicates, which
          can be set to \textit{true}, but maybe we will want different predicates depending on the context (e.g. discrete or continuous time).
    \item $T$ should be merged discrete and continuous.
    \item Need to resolve undefined with time functions.
    \item Last n-th occurence.
    \item From time to predicates that hold? What would that be useful for? Do we really want to reference predicates in other predicates?
    \item \textbf{Open question -- Channel/transmit semantics}: the transmitted value persists on the channel until the next \textit{transmit}, so \textit{receive} at any $t' \geq t$ reads the last transmitted value. It is unclear whether this is the right model for all channel types, or whether some channels should require the receiver to observe the value at the exact transmission time.
    \item \textbf{Future addition -- Time-when expression}: add $\mathit{LastTrue}_\rho(\psi, t)$ returning the last $t' \leq t$ at which $\psi(\rho, t') = \mathit{true}$, and symmetrically $\mathit{FirstTrue}_\rho(\psi, t)$. Currently only $\mathit{LastOcc}_\rho$ and $\mathit{FirstOcc}_\rho$ cover this for \texttt{Event} entities specifically.
    \item \textbf{Open question -- Synchronous channel model}: the current model is asynchronous -- \textit{transmit} and \textit{receive} are independent operations that may occur at different times. An alternative synchronous model would have \textit{transmit} directly cause \textit{received} to fire at the same time $t$, making \textit{receive} a consequence of \textit{transmit} rather than a separate operation.
\end{itemize}

\section{Examples}

\subsection{Custom Counter}

\textbf{Requirement:} After the system enters emergency mode, at each discrete time step the counter \textit{counter\_1} is incremented by 3.

\medskip
\textbf{Entities:}
\begin{itemize}
    \item $\mathit{enter\_emergency} \in E,\ t_e = \text{Event}$
    \item $\mathit{counter\_1} \in E,\ t_e = \text{Signal}$
\end{itemize}

\textbf{Constraint:}
\begin{align*}
    \mathit{Always}\Bigl(\;
        \mathit{HasHappened}_\rho(\mathit{enter\_emergency},\, \mathit{Now})
        \Rightarrow \mathit{Val}_\rho(\mathit{counter\_1},\, \mathit{Now}) = \mathit{Val}_\rho(\mathit{counter\_1},\, \mathit{Prev}(\mathit{Now})) + 3
    \;\Bigr)
\end{align*}

The antecedent holds at time $t$ iff $\mathit{enter\_emergency}$ has happened at least once up to $t$.
The consequent requires $\mathit{counter\_1}$ to equal its previous value plus 3 at every such time step.

\subsection{Req 01 -- Arithmetic recurrence with condition}

\textbf{Requirement:} The signal \textit{EngagementNoSat\_u} shall equal $1.2 \times \mathit{EngagementNoSat\_u}(k{-}1)$ when \textit{Switch\_b} is true, and \textit{Lon\_u} otherwise. The initial value is 0.

\medskip
\textbf{Entities:}
\begin{itemize}
    \item $\mathit{EngagementNoSat\_u} \in E,\ t_e = \text{Signal}$
    \item $\mathit{Switch\_b} \in E,\ t_e = \text{Signal}$
    \item $\mathit{Lon\_u} \in E,\ t_e = \text{Signal}$
\end{itemize}

\textbf{Flavour:} Discrete.

\textbf{Constraint:}
\begin{align*}
    &\mathit{Initial}\bigl(\mathit{Val}_\rho(\mathit{EngagementNoSat\_u},\, 0) = 0\bigr) \\
    \land\; &\mathit{Always}\Bigl(\\
    &\quad \bigl(\mathit{Val}_\rho(\mathit{Switch\_b},\, \mathit{Now}) = \mathit{true}\bigr) \Rightarrow {} \\
    &\qquad \mathit{Val}_\rho(\mathit{EngagementNoSat\_u},\, \mathit{Now}) = 1.2 \cdot \mathit{Val}_\rho(\mathit{EngagementNoSat\_u},\, \mathit{Prev}(\mathit{Now}))\\
    &\land\; \bigl(\mathit{Val}_\rho(\mathit{Switch\_b},\, \mathit{Now}) \neq \mathit{true}\bigr) \Rightarrow {} \\
    &\qquad \mathit{Val}_\rho(\mathit{EngagementNoSat\_u},\, \mathit{Now}) = \mathit{Val}_\rho(\mathit{Lon\_u},\, \mathit{Now})
    \;\Bigr)
\end{align*}

\subsection{Req 02 -- Store to non-volatile memory}

\textbf{Requirement:} The software shall store the maximum execution time measurement data in non-volatile memory at address \textit{MEASUREMT\_BLOCK}.

\textbf{TOTO ASI UPLNE NENI TO CO CHCEME. :(}

\medskip
\textbf{Entities:}
\begin{itemize}
    \item $\mathit{max\_exec\_time\_data} \in E,\ t_e = \text{Abstract}$
    \item $\mathit{MEASUREMT\_BLOCK} \in E,\ t_e = \text{Storage},\ \mu_e = \{\mathit{non\_volatile} \mapsto \top\}$
\end{itemize}

\textbf{Flavour:} Discrete.

\textbf{Constraint:}
\[
    \mathit{Eventually}\Bigl(
        \mathit{Val}_\rho(\mathit{MEASUREMT\_BLOCK},\, \mathit{Now}) = \mathit{Val}_\rho(\mathit{max\_exec\_time\_data},\, \mathit{Now})
    \Bigr)
\]

\subsection{Req 03 -- Ordered read then store to CPU register}

\textbf{Requirement:} The software must perform the following actions in the specified order:
1) Read the calibration constant value at the address 0xAA000018;
2) Store the calibration constant value in the DTSCON register.

\medskip
\textbf{Entities:}
\begin{itemize}
    \item $\mathit{calibration\_const} \in E,\ t_e = \text{Storage},\ \mu_e = \{\mathit{address} \mapsto \texttt{0xAA000018}\}$
    \item $\mathit{DTSCON} \in E,\ t_e = \text{Storage},\ \mu_e = \{\mathit{register} \mapsto \top\}$
\end{itemize}

\textbf{Flavour:} Discrete.

\textbf{Constraint:}
\[
    \mathit{Sequence}\Bigl(
        \mathit{Happening}(\mathit{read}(\mathit{calibration\_const}),\, \mathit{Now}),\;
        \mathit{Happening}(\mathit{written}(\mathit{DTSCON}),\, \mathit{Now})
    \Bigr)
\]

\subsection{Req 04 -- Exception vector table configuration}

\textbf{Requirement:} The software must configure IVORx registers with the addresses of Exception Vectors that compose the Exception Vector Table, according to the following table:
\begin{center}
\begin{tabular}{ll}
    \toprule
    \textbf{IVORx} & \textbf{Exception} \\
    \midrule
    IVOR0 & Critical input \\
    IVOR1 & Machine check \\
    \bottomrule
\end{tabular}
\end{center}

\medskip
\textbf{Entities:}
\begin{itemize}
    \item $\mathit{IVOR0} \in E,\ t_e = \text{Storage},\ \mu_e = \{\mathit{register} \mapsto \top\}$
    \item $\mathit{IVOR1} \in E,\ t_e = \text{Storage},\ \mu_e = \{\mathit{register} \mapsto \top\}$
    \item $\mathit{CriticalInput} \in E,\ t_e = \text{Storage}$
    \item $\mathit{MachineCheck} \in E,\ t_e = \text{Storage}$
\end{itemize}

\textbf{Flavour:} Discrete.

\textbf{Constraint:}
\[
    \mathit{Eventually}\Bigl(
        \mathit{Val}_\rho(\mathit{IVOR0},\, \mathit{Now}) = \mathit{Addr}(\mathit{CriticalInput})
        \;\land\;
        \mathit{Val}_\rho(\mathit{IVOR1},\, \mathit{Now}) = \mathit{Addr}(\mathit{MachineCheck})
    \Bigr)
\]

\subsection{Req 13 -- Bounded count of large register values}

\textbf{Requirement:} The number of values larger than 10 in the four registers A, B, C, D when they are read at the same timestep is never greater than 2.

\medskip
\textbf{Entities:}
\begin{itemize}
    \item $A \in E,\ t_e = \text{Storage},\ \mu_e = \{\mathit{register} \mapsto \top\}$
    \item $B \in E,\ t_e = \text{Storage},\ \mu_e = \{\mathit{register} \mapsto \top\}$
    \item $C \in E,\ t_e = \text{Storage},\ \mu_e = \{\mathit{register} \mapsto \top\}$
    \item $D \in E,\ t_e = \text{Storage},\ \mu_e = \{\mathit{register} \mapsto \top\}$
\end{itemize}

\textbf{Flavour:} Discrete.

\textbf{Constraint:}

Let $\mathit{big}(e) \;\triangleq\; \mathit{Val}_\rho(e,\, \mathit{Now}) > 10$ for $e \in \{A, B, C, D\}$.
``At most 2 of the four registers exceed 10'' is equivalent to ``no 3-element subset all exceed 10'', i.e.\ $\neg(\text{at least 3 exceed 10})$.
Since there are $\binom{4}{3} = 4$ such subsets, the constraint is:

\begin{align*}
    \mathit{Always}\Bigl(
        &\mathit{AllOf}\bigl(\{\,
            \mathit{Happening}(\mathit{read}(A),\, \mathit{Now}),\;
            \mathit{Happening}(\mathit{read}(B),\, \mathit{Now}),\\
        &\quad
            \mathit{Happening}(\mathit{read}(C),\, \mathit{Now}),\;
            \mathit{Happening}(\mathit{read}(D),\, \mathit{Now})
        \,\}\bigr) \\
        &\Rightarrow \mathit{Not}\Bigl(\mathit{AnyOf}\bigl(\{\\
        &\qquad \mathit{AllOf}(\{\mathit{big}(A),\, \mathit{big}(B),\, \mathit{big}(C)\}),\\
        &\qquad \mathit{AllOf}(\{\mathit{big}(A),\, \mathit{big}(B),\, \mathit{big}(D)\}),\\
        &\qquad \mathit{AllOf}(\{\mathit{big}(A),\, \mathit{big}(C),\, \mathit{big}(D)\}),\\
        &\qquad \mathit{AllOf}(\{\mathit{big}(B),\, \mathit{big}(C),\, \mathit{big}(D)\})\\
        &\}\bigr)\Bigr)
    \Bigr)
\end{align*}

The antecedent holds at time $t$ iff all four registers are read simultaneously.
The consequent rules out every possible combination of three registers all exceeding 10, which is necessary and sufficient for at most two values to be greater than 10.

\end{document}